---
title: "Machine learning project"
author: "koem32"
date: "September 12, 2014"
output: html_document
---
#Abstract

This research project aims to predict the type of activity performed by 6 subjects using data measurements from accelerometers on  the belt, forearm, arm, and dumbell. We used a training and testing set to perform this prediction and applied a random forest model to generate a predictor with an estimated out of sample error rate of less than 0.5%. As a result we believe this to be a very accurate predictor. Applying this predictor to our testing set resulted in 100% accuracy

The data set is available from the website here: http://groupware.les.inf.puc-rio.br/har. 

#Data Preprocessing

Create a file in the working directory called "machineLearningProjectKoem32" if it does not already exist and download the files into it. 
```{r}
if (!file.exists("machineLearningProjectKoem32")) {dir.create("machineLearningProjectKoem32")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileUrl, destfile = "./machineLearningProjectKoem32/pml-training.csv", method = "curl")
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl2, destfile = "./machineLearningProjectKoem32/pml-testing.csv", method = "curl")

training<-read.csv(file="machineLearningProjectKoem32/pml-training.csv",head=TRUE,sep=",")
testing<-read.csv(file="machineLearningProjectKoem32/pml-testing.csv",head=TRUE,sep=",")
```
The first step is to remove those variables in the training set which have almost 0 variance as these variables will be very weak predictors.Apply the same to the testing set

```{r, echo=FALSE}
library(caret)
nsv<-nearZeroVar(training, saveMetrics=TRUE)
omit<-which(nsv$nzv==TRUE)
training2<-training[,-omit]
testing2<-testing[,-omit]
```
Next we get rid of all column which are full of NAs in the training set and similarly remove these same columns from testing set. 
We also remove the first 6 columns (x column is just a counter and the other time variables are uniterpretable)

```{r}
training3<-training2[,apply(training2,2,function(xx) !any(is.na(xx)))] 
testing3<-testing2[,apply(training2,2,function(xx) !any(is.na(xx)))]
training4<-training3[,-(1:6)]
testing4<-testing3[,-(1:6)]
```
Next we check there are no NAs or blank fields in the training set
```{r}
sum(training4=="")
sum(is.na(training4))
```
Looks ok so we are good to go.

#Applying a prediction model

In order to estimate the out of sample error rate we create a new training and cross validation set from the training set
```{r}
set.seed(234)
inTrain <- createDataPartition(y=training4$classe, p=0.75, list=FALSE)
newTraining<-training4[inTrain,]
newTesting<-training4[-inTrain,]
```

Since we are trying to find a classification prediction, we are going to try applying random forests 
We apply the random forest algorithm to new training set and predict on the cross validation set to get an estimate of out of sample error
```{r}
library(randomForest)
set.seed(354)
modFit<-randomForest(newTraining$classe~.,data=newTraining,ntree=300,keep.forest=TRUE)
pred<-predict(modFit, newTesting)
confusionMatrix(pred,newTesting$classe)
```
Hence the estimated out of sample error rate is estimated at 1-0.9951=0.5%

When I plot the MSE of each classe I note how I could have gotten a pretty good predictor with just 50 or 60 trees as the MSE drops quickly after the first 50 trees.
```{r}
plot(modFit,type="l")
legend("topright", legend=unique(newTraining$classe) )
```
Now let's apply this model to the original testing set
```{r}
answers<-predict(modFit, testing4[,-53])
answers
```
Finally I can download the answers into txt files in a folder names machineLearningAnswers into the working directory so as to submit them for this project
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
if (!file.exists("machineLearningAnswers")) {dir.create("machineLearningAnswers")}
setwd("./machineLearningAnswers")
pml_write_files(answers)
```